---
title: "Hands-on Exercise 1"
date: "8 January 2024"
format: html
execute: 
  echo: true #shows the code
  eval: true #shows the outcomes of the code
  warning: false #does not show the warnings
editor: visual
background-color: lightgrey;
font-family:  Palatino Linotype;
---

# Hands-on Exercise 1

## 1. Getting Started

### Installing and Launching R Packages

The code chunk below uses the `p_load()` function of the **pacman** package to check if the tidyverse packages are installed in the computer. If yes, they are then launched in R.

```{r}
pacman::p_load(tidyverse)
```

### Importing the Data

The data set for this hands-on exercise is imported into the R environment using the `read_csv()` function of the **readr** package and stored as the R object, `exam_data`.

The tibble data frame, `exam_data`, has 7 columns and 322 rows.

```{r}
exam_data = read_csv("data/Exam_data.csv")
```

### Plotting a Simple Bar Chart

The `ggplot()` and `geom_bar()` functions in the **ggplot2** package are used to plot a bar chart showing the breakdown of the data by the "RACE" variable.

From the bar chart, we can see that most of the students are Chinese, followed by Malay. A minority of students are Indian or other races.

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```
