---
title: "Hands-on Exercise 7C"
date: "26 February 2024"
date-modified: "last-modified" #allows for updating to the latest date
format: html
execute: 
  echo: true #shows the code
  eval: true #shows the outcomes of the code
  warning: false #does not show the warnings
editor: visual
background-color: lightgrey;
font-family:  Palatino Linotype;
---

# 23 Analytical Mapping

## 23.1 Overview and Learning Outcomes

This hands-on exercise is based on [Chapter 9](https://r4va.netlify.app/chap09) of the **R for Visual Analytics** book.

Visualising distribution is not new in statistical analysis. [Chapter 1](https://r4va.netlify.app/chap01) covered some of the popular statistical graphic methods for visualising distribution. They are histogram, probability density curve (pdf), boxplot, notch plot, and violin plot, using the [**ggplot2**](https://ggplot2.tidyverse.org/) package. Chapter 9 will cover two relatively new statistical graphic methods for visualising distribution. They are ridgeline plot and raincloud plot using the **ggplot2** package and its extensions.

## 23.2 Getting Started

### 23.2.1 Installing and Loading Required Libraries

In this hands-on exercise, the following R packages are used:

-   [**tidyverse**](https://www.tidyverse.org/) (i.e. readr, tidyr, dplyr) for performing data science tasks such as importing, tidying, and wrangling data;

-   [**ggthemes**](https://ggplot2.tidyverse.org/reference/ggtheme.html) for extra themes, geoms, and scales for ggplot2;

-   [**ggridges**](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html) for creating ridgeline plots; and

-   [**ggdist**](https://mjskay.github.io/ggdist/) for visualising distributions and uncertainty.

The code chunk below uses the `p_load()` function in the **pacman** package to check if the packages are installed. If yes, they are then loaded into the R environment. If no, they are installed, then loaded into the R environment.

```{r}
#| eval: false
pacman::p_load(tidyverse, ggthemes,
               ggridges, ggdist)
```

### 23.2.2 Importing Data

The dataset for this hands-on exercise is imported into the R environment using the `read_csv()` function in the **readr** package and stored as the R object, `exam`.

```{r}
#| eval: false
exam = read_csv("data/Exam_data.csv")
```

The tibble data frame, `exam_data`, has 7 columns and 322 rows.

-   It consists of the year-end examination grades of a cohort of 322 Primary 3 students from a local school.

-   The 7 variables/attributes are:

    -   Categorical: ID, CLASS, GENDER, and RACE.

    -   Continuous: MATHS, ENGLISH, and SCIENCE.

## 23.3 Basic Choropleth Mapping

### 23.3.1 Visualising Distribution of Non-functional Water Point

xxx

## 23.4 Choropleth Map for Rates

xxx

### 23.4.1 Deriving Proportion of Functional Water Points and Non-functional Water Points

xxx

### 23.4.2 Plotting Map of Rate

xxx

## 23.5 Extreme Value Maps

xxx

### 23.5.1 Percentile Map

xxx

### 23.5.1 Percentile Map

xxx

[**\~\~\~ End of Hands-on Exercise 7C \~\~\~**]{.smallcaps}
