---
title: "Hands-on Exercise 4C"
date: "28 January 2024"
date-modified: "last-modified" #allows for updating to the latest date
format: html
execute: 
  echo: true #shows the code
  eval: true #shows the outcomes of the code
  warning: false #does not show the warnings
editor: visual
background-color: lightgrey;
font-family:  Palatino Linotype;
---

# 11 Visualising Uncertainty

## 11.1 Overview and Learning Outcomes

This hands-on exercise is based on [Chapter 3](https://r4va.netlify.app/chap03) of the **R for Visual Analytics** book.

The learning outcome is to create interactive data visualisation using functions in the [**ggiraph**](https://davidgohel.github.io/ggiraph/) and [**plotlyr**](https://plotly.com/r/) package.

## 11.2 Getting Started

### 11.2.1 Installing and Loading Required Libraries

In this hands-on exercise, the following R packages are used:

-   [**tidyverse**](https://www.tidyverse.org/) (i.e. readr, tidyr, dplyr) for performing data science tasks such as importing, tidying, and wrangling data;

-   [**patchwork**](https://patchwork.data-imaginist.com/) for preparing composite figures created using ggplot2;

-   [**DT**](https://rstudio.github.io/DT/) for interfacing with JavaScript library DataTables that create interactive tables on html pages;

-   [**ggiraph**](https://davidgohel.github.io/ggiraph/) for making ggplot graphics interactive;

-   [**plotlyr**](https://plotly.com/r/) for plotting interactive statistical graphs; and

-   [**crosstalk**](https://rstudio.github.io/crosstalk/) for inter-widget interactivity for html widgets.

The code chunk below uses the `p_load()` function in the **pacman** package to check if the packages are installed. If yes, they are then loaded into the R environment. If no, they are installed, then loaded into the R environment.

```{r}
pacman::p_load(tidyverse, patchwork,
               DT, ggiraph, plotly,
               crosstalk)
```

### 11.2.2 Importing Data

The dataset for this hands-on exercise is imported into the R environment using the `read_csv()` function in the **readr** package and stored as the R object, `exam_data`.

```{r}
exam_data = read_csv("data/Exam_data.csv")
```

The tibble data frame, `exam_data`, has 7 columns and 322 rows.

-   It consists of the year-end examination grades of a cohort of 322 Primary 3 students from a local school.

-   The 7 variables/attributes are:

    -   Categorical: ID, CLASS, GENDER, and RACE.

    -   Continuous: MATHS, ENGLISH, and SCIENCE.

## 11.3 Visualising the Uncertainty of Point Estimates: ggplot2 Methods

### 11.3.1 Plotting Standard Error Bars of Point Estimates

xxx

### 11.3.2 Plotting Confidence Interval of Point Estimates

xxx

### 11.3.3 Visualising the Uncertainty of Point Estimates with Interactive Error Bars

xxx

## 11.4 Visualising Uncertainty: ggdist Package

xxx

### 11.4.1 **Visualising the Uncertainty of Point Estimates: ggdist Methods**

xxx

### 11.4.2 Visualising the Uncertainty of Point Estimates: ggdist Methods

xxx

### 11.4.3 Visualising the Uncertainty of Point Estimates: ggdist Methods

xxx

## 11.5 **Visualising Uncertainty with Hypothetical Outcome Plots (HOPs)**

xxx

[**\~\~\~ End of Hands-on Exercise 4C \~\~\~**]{.smallcaps}
