---
title: "Hands-on Exercise 7A"
date: "25 February 2024"
date-modified: "last-modified" #allows for updating to the latest date
format: html
execute: 
  echo: true #shows the code
  eval: true #shows the outcomes of the code
  warning: false #does not show the warnings
editor: visual
background-color: lightgrey;
font-family:  Palatino Linotype;
---

# 21 Choropleth Mapping

## 21.1 Overview and Learning Outcomes

This hands-on exercise is based on [Chapter 9](https://r4va.netlify.app/chap09) of the **R for Visual Analytics** book.

Visualising distribution is not new in statistical analysis. [Chapter 1](https://r4va.netlify.app/chap01) covered some of the popular statistical graphic methods for visualising distribution. They are histogram, probability density curve (pdf), boxplot, notch plot, and violin plot, using the [**ggplot2**](https://ggplot2.tidyverse.org/) package. Chapter 9 will cover two relatively new statistical graphic methods for visualising distribution. They are ridgeline plot and raincloud plot using the **ggplot2** package and its extensions.

## 21.2 Getting Started

### 21.2.1 Installing and Loading Required Libraries

In this hands-on exercise, the following R packages are used:

-   [**tidyverse**](https://www.tidyverse.org/) (i.e. readr, tidyr, dplyr) for performing data science tasks such as importing, tidying, and wrangling data;

-   [**ggthemes**](https://ggplot2.tidyverse.org/reference/ggtheme.html) for extra themes, geoms, and scales for ggplot2;

-   [**ggridges**](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html) for creating ridgeline plots; and

-   [**ggdist**](https://mjskay.github.io/ggdist/) for visualising distributions and uncertainty.

The code chunk below uses the `p_load()` function in the **pacman** package to check if the packages are installed. If yes, they are then loaded into the R environment. If no, they are installed, then loaded into the R environment.

```{r}
#| eval: false

pacman::p_load(tidyverse, ggthemes,
               ggridges, ggdist)
```

### 21.2.2 Importing and Preparing Aspatial Data

The dataset for this hands-on exercise is imported into the R environment using the `read_csv()` function in the **readr** package and stored as the R object, `exam`.

```{r}
#| eval: false

exam = read_csv("data/Exam_data.csv")
```

The tibble data frame, `exam_data`, has 7 columns and 322 rows.

-   It consists of the year-end examination grades of a cohort of 322 Primary 3 students from a local school.

-   The 7 variables/attributes are:

    -   Categorical: ID, CLASS, GENDER, and RACE.

    -   Continuous: MATHS, ENGLISH, and SCIENCE.

### 21.2.3 Importing and Preparing Geospatial Data

xxx

### 21.2.4 Joining Aspatial and Geospatial Data

xxx

## 21.3 Plotting Choropleth Maps Using tmap

xxx

### 21.3.1 Plotting Choropleth Map: qtm()

xxx

### 21.3.2 Plotting Choropleth Map: tmap Package

xxx

#### 21.3.2.1 Drawing Base Map

xxx

#### 21.3.2.2 Drawing Choropleth Map: tm_polygons()

xxx

#### 21.3.2.3 Drawing Choropleth Map: tm_fill() and tm_border()

xxx

#### 21.3.2.1 Drawing Base Map

xxx

### 21.3.3 Data Classification Methods of tmap

xxx

xxx

### 21.3.4 Colour Scheme

xxx

### 21.3.5 Map Layouts

xxx

#### 21.3.5.1 Legend

xxx

#### 21.3.5.2 Style

xxx

#### 21.3.5.3 Cartographic Furniture

xxx

### 21.3.6 Drawing Small Multiple Choropleth Maps

xxx

#### 21.3.6.1 By Assigning Multiple Values to Aesthetic Argument

xxx

#### 21.3.6.2 By Defining A Group-by Variable in tm_facets()

xxx

#### 21.3.6.3 By Creating Multiple Stand-alone Maps with tmap_arrange()

xxx

### 21.3.7 Mapping Spatial Object Meeting Selection Criterion

xxx

## 21.4 References

xxx

xxx

[**\~\~\~ End of Hands-on Exercise 7A \~\~\~**]{.smallcaps}
