---
title: "In-class Exercise 1: Now You See It!"
date: "13 January 2024"
date-modified: "last-modified" #allows for updating to the latest date
format: html
execute: 
  echo: true #shows the code
  eval: true #shows the outcomes of the code
  warning: false #does not show the warnings
editor: visual
background-color: lightgrey;
font-family:  Palatino Linotype;
---

***Note***: Updated on 18 January 2024 to make editorial edits.

# Now You See It!

## 1 Overview and Learning Outcomes

This is the in-class exercise 1.

The learning outcomes are to load the relevant R packages and import the required dataset into the environment.

## 2 Loading R Packages

In this in-class exercise, two R packages will be used. They are:

-   [**tidyverse**](https://www.tidyverse.org/) (i.e. readr, tidyr, dplyr) for performing data science tasks such as importing, tidying, and wrangling data; and

-   [**haven**](https://haven.tidyverse.org/) for importing SAS files.

The code chunk below uses the `p_load()` function in the **pacman** package to check if the packages are installed in the computer. If yes, they are then loaded into the R environment. If no, they are installed, then loaded into the R environment.

```{r}
pacman::p_load(tidyverse, haven)
```

## 3 Importing Data

The dataset used in this in-class exercise is the PISA 2022 student questionnaire data file, which is in the SAS file format. It is imported into the R environment using the [`read_sas()`](https://haven.tidyverse.org/reference/read_sas.html) function in the [**haven**](https://haven.tidyverse.org/) package and stored as the R object, `stu_qqq`.

```{r}
#| eval: false
stu_qqq = read_sas("data/cy08msp_stu_qqq.sas7bdat")
```

The tibble data frame, `stu_qqq`, has 1,279 columns (variables) and 613,744 rows (observations).

There are 6,606 rows with the country code "SGP", which represents Singapore. The [`filter()`](https://dplyr.tidyverse.org/reference/filter.html) function in the [**dplyr**](https://dplyr.tidyverse.org/) package is used to obtain these rows, and stored as the R object, `stu_qqq_SG`.

```{r}
#| eval: false
stu_qqq_SG = stu_qqq %>% filter(CNT == "SGP")
```

The tibble data frame, `stu_qqq_SG`, is then saved in the rds file format and imported into the R environment.

```{r}
#| eval: false
write_rds(stu_qqq_SG, "data/stu_qqq_SG.rds")
```

```{r}
stu_qqq_SG = read_rds("data/stu_qqq_SG.rds")
```

[**\~\~\~ End of In-class Exercise 1 \~\~\~**]{.smallcaps}
